Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-06-11 manuscript/manuscript.tex
Add Helmut to author list
Add Dan's recent MPI parallelization to tutorial and bibliography

\author[1\authfn{1}]{Franz Waibl}
\author[1\authfn{1}]{Valentin J Hoerschinger}
\author[2]{Vjay Molino}
\author[1]{Monica L Fern{\'a}ndez-Quintero}
%\author[1,2]{Firstname Middlename Familyname}
%\author[2]{Firstname Initials Surname}
\author[2]{Steven Ramsey}
\author[4]{Daniel R Roe}
\author[1*]{Klaus R Liedl}
\author[3*]{Michael K Gilson}
\author[2*]{Tom Kurtzman}

@@

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and show how to avoid technical difficulties.
We assume familiarity with molecular dynamics simulations as well as the AmberTools package.

@@

GIST is an implementation of IST on a three-dimensional grid. It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on a space filling grid, rather than high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.

Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-states approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
%Here you would explain what problem you are tackling and briefly motivate your work.
%
%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
%
%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications. Usually, this would be one of your figures, but it must also be uploaded separately upon article submission. We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
%
%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
The user should be able to run MD simulations and analyze the resulting trajectories.
Previous experience of GIST is not necessary, although we recommend reading the articles referenced in section~\ref{sec:theory}.

@@

The Amber simulation engine is used in this tutorial to run the MD simulations.
While Amber is free for non-commercial use, it might not always be available in industrial settings.
In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
%We should actually test that!
Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. For further reference, the influence of compression on energy calculations was discussed in a recent publication \cite{Roe2022-compression}.

@@

\paragraph{Identify and define the region of interest}
In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
This region is represented by a rectangular grid with given position, size, and voxel spacing.
The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
While integrals of GIST quantities over the grid are independent of the voxel size, values for each voxel converge slower at high resolutions.

@@

To use OpenMP or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP} or \inlinecode{cpptraj.cuda} is often used. Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to  GIST.
The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
A full list is provided at the end of this tutorial or through the \software{cpptraj} manual.
The most important options are summarized in the following table:

@@

Due to the varying sizes of the boxes, the reference energies used in this context may not align with those specified in the Amber manual. Generally, it is recommended that the size of the reference water box closely matches that of the production system. Therefore, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small. While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the energy per water molecule ($E_{ww, norm}$) for water boxes. These discrepancies, when summing over large grid regions, have the potential to introduce significant errors due to incorrect referencing.

@@

The output is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
Energy and entropy quantities are reported in two different ways.
The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
The \inlinecode{\_norm} quantities are instead normalized by the solvent number density in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
To convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel $k$ to a \inlinecode{\_dens} quantity $X^k_{dens}$, the following equation can be applied:
\begin{equation}
	X^k_{dens} = \frac{X^k_{norm}}{V^k} \cdot \frac{N^k_{solvent}}{N_{frames}}
\end{equation}
with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the number of solvent molecules in $k$ over all simulation frames, and $N_{frames}$, the number of frames.

@@

In the first part, we show how to run GIST calculations for small molecules and interpret the output.
Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST. In addition, we discuss all steps in further detail at the end.

@@

Now, you can finally run the actual GIST analysis.
Center the molecule as above, to verify that its position matches the grid parameters you decided on.
We recommend using either the PME or GPU implementation of GIST.
For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc
autoimage !(:WAT) origin
# optionally remove the "pme" option
gist griddim 100 100 100 out gist.dat \
refdens 0.03287 pme
\end{lstlisting}
\subsubsection{Working with the GIST results}
Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the grid is produced. 
Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.

@@

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}

@@

This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
\subsection{Applying GIST for the Streptavidin-Biotin complex}
In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline potential pitfalls. We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
\subsubsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}[style=bash]
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

@@

Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can ensure that the monomer conformations match the complex.

@@

\subsubsection{GIST analysis of Streptavidin-Biotin}
For easier post-processing, here we use grids that contain the whole molecule.
Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
Furthermore, we recommend that you center the molecule to the coordinate origin.
Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to fully capture the perturbation of the solvent by the solute.
Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.

For the GIST analysis, we recommend using either the PME or GPU implementation of GIST.
For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.

@@

\subsection{PME implementation}
In Reference~\cite{Chen2021}, a PME implementation of the GIST energy was presented.
For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an MD simulation.

To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.
\subsection{GIST with non-water solvents}
In References~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.

@@

The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
To compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.

@@

\subsection{Installation}
GIST calculations in cpptraj can be sped up through various levels of 
parallelization.
If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
Furthermore, if cpptraj was compiled with OpenMP support, the entropy calculation will be parallelized.

>>>>

\author[1\authfn{1}]{Franz Waibl}
\author[1\authfn{1}]{Valentin J Hoerschinger}
\author[2]{Vjay Molino}
\author[1]{Monica L Fern{\'a}ndez-Quintero}
%\author[1,2]{Firstname Middlename Familyname}
%\author[2]{Firstname Initials Surname}
\author[4]{Daniel R Roe}
\author[1*]{Klaus R Liedl}
\author[3*]{Michael K Gilson}
\author[2*]{Tom Kurtzman}

@@

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis at the example of the streptavidin-biotin complex.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and show how to avoid technical difficulties.
We assume familiarity with molecular dynamics simulations as well as the AmberTools package.

@@

GIST is an implementation of IST on a three-dimensional grid. It can be seen as a more general version of WaterMap \cite{Abel2008-watermap,Young2007-watermap} that provides solvation thermodynamics on the entire grid, rather than high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.

Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
%Here you would explain what problem you are tackling and briefly motivate your work.
%
%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
%
%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications. Usually, this would be one of your figures, but it must also be uploaded separately upon article submission. We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
%
%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
The user should be able to run MD simulations and analyze the resulting trajectories.
Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.

@@

The Amber simulation engine is used in this tutorial to run the MD simulations.
While Amber is free for non-commercial use, it might not always be available in industry settings.
In that case, other MD engines can be used as long as topologies and trajectories can be produced in a format supported by \software{cpptraj}.
If using GROMACS \cite{Abraham2015-gromacs,Pronk2013-gromacs}, we propose to prepare the topology with AmberTools and convert it using \software{acpype.py} \cite{Sousa_da_Silva2012-acpype}.
%We should actually test that!
Also note that compressed trajectory formats such as \inlinecode{xtc} might bias the entropy calculation due to the loss of precision for the atom positions. For further reference, the influence of compression on energy calculations was discussed in a recent publication \cite{Roe2022-compression}.

@@

\paragraph{Identify and define the region of interest}
In order to run a GIST analysis, one must define the region where solvent properties will be calculated.
This region is represented by a rectangular grid with given position, size, and voxel spacing.
The default value of \SI{0.5}{\angstrom} provides reasonable discretization for many use-cases.
While integrals of GIST quantities over the grid are independent of the voxel size, individual values converge slower at high resolutions.

@@

To use OpenMP, MPI or GPU-accelerated (using CUDA) versions of cpptraj, a different executable name such as \inlinecode{cpptraj.OMP}, \inlinecode{cpptraj.MPI} or \inlinecode{cpptraj.cuda} is often used.
Therefore, make sure to use the right compilation and executable, to get full benefit of the various improvements to GIST.
The \inlinecode{gist} action has a large number of options to modify the GIST calculation or its output.
A full list is provided at the end of this tutorial or through the \software{cpptraj} manual.
The most important options are summarized in the following table:

@@

Due to the varying sizes of the boxes, the reference energies used in this context may not align with those specified in the Amber manual. Generally, it is recommended that the size of the reference water box closely matches that of the production system. Therefore, it is advisable to conduct your own reference calculations, especially for systems that are either very large or very small. While the impact on density is not as pronounced, deviations from the specified settings can lead to differences of up to 0.005 kcal/mol in the $E_{ww, norm}$ value for water boxes. These discrepancies, when summing over large grid regions, have the potential to introduce significant errors due to incorrect referencing.

@@

The output is organized by voxel and sorted according to their x, y, and z coordinates, describing the center point of the voxel.
Energy and entropy quantities are reported in two different ways.
The \inlinecode{\_dens} quantities are normalized by the voxel volume and can be used to compute integrals.
The \inlinecode{\_norm} quantities are instead normalized by the number of solvents in the respective voxel averaged over all considered frames, and are useful to visualize local water properties.
To convert from a \inlinecode{\_norm} quantity $X^k_{norm}$ in voxel $k$ to a \inlinecode{\_dens} quantity $X^k_{dens}$, the following equation can be applied:
\begin{equation}
	X^k_{dens} = \frac{X^k_{norm}}{V^k} \cdot \frac{N^k_{solvent}}{N_{frames}}
\end{equation}
with $V^k$, the volume of voxel $k$, $N^k_{solvent}$, the number of solvent molecules in $k$ over all simulation frames, and $N_{frames}$, the number of frames.

@@

In the first part, we show how to run GIST calculations for small molecules and interpret the output.
Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST. In addition, we discuss all steps in further detail at the end.

@@

Now, you can finally run the actual GIST analysis.
Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
We recommend using either the GPU or MPI-accelerated PME implementation of GIST.
While the GPU version is faster when limited CPU resources are available, the PME implementation matches the energy of the Amber MD engine more closely.
More recently, MPI support was added for both the energy and entropy calculation\cite{Roe2023-mpi-gist}. 
The MPI implementation scales well over multiple CPU cores and can be combined with either the GPU or PME implementations for maximum efficiency. 
A speed comparison of the various parallelization implementations is shown in \cite{Roe2023-mpi-gist}.

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc
autoimage !(:WAT) origin
# optionally remove the "pme" option
gist griddim 100 100 100 out gist.dat \
refdens 0.03287 pme
\end{lstlisting}
\subsubsection{Working with the GIST results}
Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.

@@

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}

@@

This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
\subsection{Applying GIST for the Streptavidin-Biotin complex}
In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls. We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
\subsubsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}[style=bash]
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

@@

Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can make sure that the monomer conformations match the complex.

@@

\subsubsection{GIST analysis of Streptavidin-Biotin}
For easier post-processing, here we use grids that contain the whole molecule.
Smaller grids encompassing only the binding site could be used for streptavidin and the complex, but this would complicate the analysis.
Furthermore, we recommend that you center the molecule to the coordinate origin.
Calculate the GIST grid parameters as outlined in the biotin section for the complex and streptavidin and run the GIST analysis.
For larger systems like whole proteins you can usually get away with a smaller buffer region compared to small molecules.
To test whether your grid is large enough look at solvent properties close to the grid edges, e.g. the solvent-solvent energies.
If these correspond to the respective bulk properties, it can be assumed that your grid is large enough to capture all perturbations of the solute on the solvent.
Note that a different box-size or number of solvent molecules can skew the bulk properties slightly compared to the reference values provided above.
In that case, checking for convergence of the solvent properties near the grid borders is sufficient but you might want to modify your reference values according to your results then as described in the next section.

@@

\subsection{PME implementation}
In Reference~\cite{Chen2021}, a PME implementation of the GIST energy was presented.
For typical systems, this implementation is slightly slower than GPU-GIST, but much faster than the original CPU code.
Furthermore, PME-GIST offers the best agreement between GIST energies and those observed in an classical MD simulation.

To run PME-GIST, the \inlinecode{pme} flag to \inlinecode{gist} can be used in \software{cpptraj}.
The output will contain two additional columns: \inlinecode{PME\_dens} and \inlinecode{PME\_norm}.
They contain the potential of solute and solvent evaluated at the solvent positions and divided by two.
Furthermore, the \inlinecode{Eww} and \inlinecode{Esw} columns are also computed using PME, and are reported as usual.
For typical use cases, we recommend using \inlinecode{Eww} and \inlinecode{Esw} over the \inlinecode{PME} column.

\subsection{MPI parallelization}
The most recent addition to GIST is the MPI parallelization of both energy and entropy calculations\cite{Roe2023-mpi-gist}.
Since the MPI parallelization is orthogonal to the other improvements, it can be used with both PME or GPU accelerated GIST.
\subsection{GIST with non-water solvents}
In references~\cite{Kraml2020,Kamenik2020-gist-macrocycles,Waibl2022-gist-solvents}, an extension of GIST towards solvents other than water is described.

@@

The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
Note that the calculation of the reference value is complicated by the introduction of additional solvents.
For an overview of how to correctly handle such calculations, refer to reference \cite{Waibl2021-gist-salt}.
Additionally, to compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.

@@

\subsection{Installation}
GIST calculations in cpptraj can be sped up through various levels of 
parallelization.
If cpptraj was compiled with the '-cuda' configure flag set, the GPU accelerated version of GIST is used automatically. 
In addition, OpenMP and MPI parallelization are available if cpptraj was compiled the necessary options.
